---
title: "Logistic Regression"
teaching: 60
exercises: 60
source: Rmd
---

```{r, include=FALSE}
## Limit printing output to max. 25 lines to avoid endless scrolling
options(max.print = 25)
```

::::::::::::::::::::::::::::::::::::::: objectives

- To be able to construct regression models for a binary outcome
- To be able to calculate predicted variables and residuals

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::: questions

- How can I identify factors for antibiotic resistance?
- How can I check the validity of model?

::::::::::::::::::::::::::::::::::::::::::::::::::

## Content

-   Exploratory Data Analysis
-   Model Creation and Estimation
-   Reporting the Logistic Regression Results

## Data

```{r libraries, message=FALSE, warning=FALSE}
# We will need these libraries and this data later.
library(aod)
library(broom)
library(ggplot2)
library(lubridate)
library(odds.n.ends)
library(tidyverse)

amrData <- read.csv("data/dig_health_hub_amr_v2.csv")
```
The data used in this episode was provided by Simon Thelwall from the UKHSA.
It has been created to represent the sort of data that might be obtained from the Second Generation Surveillance System (SGSS).
The data has 1000 rows of 13 variables.

## Exploratory Data Analysis

We can preview our data by using 'head':

```{r}
head(amrData)
```
We can also request information about variable names and data types:
```{r}
sapply(amrData, class)
```

We can see that dates are currently stored as the char data type. We also do not know the age of the participant when the sample was taken.

```{r}
# Calculate age (in years) as of their last birthday and add as an additional variable to our data.
# The %--% and %/% are synax specific to lubridate.
# In the first part we are asking it to find the difference between the two dates.
# We are then rounding down to the nearest year.

amrData <- amrData %>%
  mutate(
    age_years_sd = (dob %--% spec_date) %/% years(1)
  )
```

We can also convert 'spec_date', the date the specimen was taken from text to a date:
```{r}
# Convert char to date and store as additional variable
amrData <- amrData %>%
  mutate(
    spec_date_YMD = as.Date(amrData$spec_date)
  )
```
We can use a histogram to explore the age distribution of the participants:

```{r}
# histogram of age
ageHisto <- amrData %>%
  ggplot(aes(x = age_years_sd)) +
  geom_histogram(bins = 10, color = "white") +
  theme_minimal(base_size = 14, base_family = "sans") +
  labs(x = "Age when specimen taken (years)", y = "Frequency")
ageHisto
```
We can also look at where the specimens were processed:
```{r}
xtabs(~trst_cd, data = amrData)
```

and for which organism:
```{r}
xtabs(~organism, data = amrData)
```

In addition, we can use cross-tabulation to identify if the specimen indicated resistance to one or more of Coamoxiclav, Gentamicin and Ciprofloxacin for the participants:

```{r}
xtabs(~ coamox + cipro + gentam, data = amrData)
```
Coamoxiclav appears to have the highest individual indication of resistance. We will explore indicators to Coamoxiclav first.

##  Model Creation and Estimation

As the dependent variable we want to explore is binary (0,1), we will use a binomial generalised linear model.


```{r}
coamox_logit <- glm(coamox ~ age_years_sd + sex_male, data = amrData, family = "binomial")

summary(coamox_logit)
```
In this initial model focusing on demographic indicators, age at time of sample being taken seems to be statistically significant but, whether or not the participant is male both does not.

age_years_sd: For every unit increase in age_years_sd the log-odds of Coamoxiclav resistance increase by 0.0404280.

sex_male: The difference in the log-odds of Coamoxiclav resistance between males and non-males is 0.196694.

Older participants have higher log-odds of Coamoxicalv resistance.

::::::::::::::::::::::::::::::::::::::  callout

If both age_years_sd and sex_male had indicated statistical significance we would want to check whether 
these were significant independently or if there was a relationship between the two variables.

We can check to see that our indicators sex_male and age_years_sd are independent:
```{r}
# check VIF for no perfect multicollinearity assumption
car::vif(coamox_logit)
```
We can also check the linearity of the variable age_years_sd

```{r}
# make a variable of the logit of the predicted values
logit.use <- log(coamox_logit$fitted.values / (1 - coamox_logit$fitted.values))

# make a small data frame with the logit variable and the age predictor
linearity_data <- data.frame(logit.use, age = coamox_logit$model$age_years_sd)

# create a plot with linear and actual relationships shown
ggplot(linearity_data, aes(x = age, y = logit.use)) +
  geom_point(aes(size = "Observation"), color = "blue", alpha = .6) +
  geom_smooth(aes(color = "Loess curve"), se = FALSE) +
  geom_smooth(aes(color = "linear"), method = lm, se = FALSE) +
  theme_minimal(base_size = 14, base_family = "serif") +
  labs(x = "Age in years on sample date", y = "Log-odds of coamox resistance predicted probability") +
  scale_color_manual(name = "Type of fit line", values = c("red", "black")) +
  scale_size_manual(values = 1.5, name = "")
```

::::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::  challenge

### Challenge 1

Update the model coamox_logit to include had_surgery_past_yr as an independent variable.
What is the log-odds reported and is it statistically significant?

:::::::::::::::  solution

### Challenge 1

You may choose to create a new glm:

```{r}
coamox_surg_logit <- glm(coamox ~ age_years_sd + sex_male + had_surgery_past_yr, data = amrData, family = "binomial")

summary(coamox_surg_logit)
```
had_surgery_past_yr: The difference in the log-odds of Coamoxiclav resistance between those who have had surgery in the past year
and those who have not is -0.082709. It is not statistically significant.

You may also want to check for multicollinearity:

```{r}
car::vif(coamox_surg_logit)
```
As the value of GVIF is lower than 4, it suggests that the assumption of independence between the variables is held.

:::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::::::::::::::::


### Incorporating a multi-level factor

We are going to incorporate ethnicity into our model. We first need to conver this to a factor, to make sure it is
interpreted as a categorical variable.

```{r}
amrData$ethnicity <- factor(amrData$ethnicity)

coamox_surg_ethnicity_logit <- glm(
  coamox ~ age_years_sd + sex_male + had_surgery_past_yr + ethnicity,
  data = amrData,
  family = "binomial"
)

summary(coamox_surg_ethnicity_logit)
```
The ethnicity category Asian, Asian British or Asian Welsh has been used as the reference group.
Estimates for the other categories are reported in relation to this group.

We can use the Wald test to explore the overall effect of ethnicity. We will use the `wald.test` function of the `aod` library. The order in which the coefficients are given in the table of coefficients is the same as the order of the terms in the model. This is important because the `wald.test` function refers to the coefficients by their order in the model. 
`b` supplies the coefficients, while `Sigma` supplies the variance covariance matrix of the error terms, finally `Terms` tells R which terms in the model are to be tested, in this case, terms 5 to 8, which correspond to the different ehtnicities.

```{r}
wald.test(b = coef(coamox_surg_ethnicity_logit), Sigma = vcov(coamox_surg_ethnicity_logit), Terms = 5:8)
```

The chi-squared test statistic of 3, with 4 degrees of freedom is associated with a p-value of 0.56 indicating that the overall effect of ethnicity is not statistically significant.


## Reporting the Logistic Regression Results

For our model `coamox_surg_ethnicity_logit` there are various things that we can report, and different functions and packages that can be used.

We can report the confidence intervals using profiled log-likelihood or using standard errors for our model.

```{r}
# CIs using log-likelihood
confint(coamox_surg_ethnicity_logit)

## CIs using standard errors
confint.default(coamox_surg_ethnicity_logit)
```

Alternatively we may be interested in the odds-ratio:

```{r}
## odds ratios only
exp(coef(coamox_surg_ethnicity_logit))

# including confidence intervals
exp(cbind(OR = coef(coamox_surg_ethnicity_logit), confint(coamox_surg_ethnicity_logit)))
```
These can also be presented using broom:

```{r}
# as Log-Odds
tidy(coamox_surg_ethnicity_logit)

# as ORs
tidy(coamox_surg_ethnicity_logit, exp = TRUE)
```

The odd.n.ends package provides a wide range of reporting tools

```{r}
# Odds ratios and confidence intervals
coamox_surg_ethnicity_logitOR <- odds.n.ends::odds.n.ends(coamox_surg_ethnicity_logit)
coamox_surg_ethnicity_logitCI <- coamox_surg_ethnicity_logitOR$`Predictor odds ratios and 95% CI`
coamox_surg_ethnicity_logitCI

# model fit
modfit <- coamox_surg_ethnicity_logitOR$`Count R-squared (model fit): percent correctly predicted`
modfit

# Other model statistics
odds.n.ends::odds.n.ends(coamox_surg_ethnicity_logit)
```
As you can see we have a range of probabilities, odds ratios and log-odds. We need to be aware of which we are referring to.
One way is to look at the range of the estimates. Probabilities always have a range from zero to 1. Logit units generally range from about -4 to +4, with zero meaning an equal probability of no event or the event outcome occurring. Odds ratios can range from very small (but positive) numbers to very large positive numbers.

These odds ratios versions of the estimates are more easily interpreted than logit scores. Odds ratios of less than one means that an increase in that predictor makes the outcome less likely to occur, and an odds ratio of greater than one means that an increase in that predictor makes the outcome more likely to occur.

::::::::::::::::::::::::::::::::::::::: discussion

## Discussion

* What other indicators could we have included in our model?
* What question(s) would they help us to answer?

::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::  challenge

## Challenge 3

We have explored potential indicators of Coamoxiclav resistance.

Explore potential indicators of either Gentamicin or Ciprofloxacin resistance.


::::::::::::::::::::::::::::::::::::::::::::::::::

:::::::::::::::::::::::::::::::::::::::  keypoints

* Logistic regression models the log-odds of an event as a linear combination of 
one or more independent variables.

* Binary logistic regression, where  a single binary dependent variable, 
coded by an indicator variable, where the two values are labeled "0" and "1", 
can be used to model the probability of a certain class or event taking place. 
In these examples, antimicrobial resistance to a particular antibiotic.

::::::::::::::::::::::::::::::::::::::::::::::::::
